---
title: Using integrated multispecies occupancy models to map co-occurrence between
  bottlenose dolphins and fisheries in the Gulf of Lion, French Mediterranean Sea.
author: "Valentin Lauret, Hélène Labach, Léa David, Matthieu Authier, Olivier Gimenez"
subtitle: R codes - Comparing multi- and single-species integrated occupancy models
output:
  pdf_document: default
  word_document: default
  bookdown::pdf_document2:
    extra_dependencies: float
    highlight: tango
geometry: margin=0.5cm
toc: no
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval = FALSE, 
                      echo = TRUE,
                      results = 'hide', 
                      dpi = 300, 
                      cache = TRUE)
```

# Load required packages

```{r}
library(tidyverse)
library(sf)
library(nimble)
```

# Load data

```{r ,  message=FALSE, warning=FALSE}
load(here("Rcodes/IMSOdata.rdata"))
```

# Integrated single-species Occupancy

Write an integrated occupancy model for a single-species

## BUGS model

```{r eval = FALSE, include=TRUE }
intOcc <- nimbleCode({
  
  ## state process
  for(j in 1:nsite){
    z[j] ~ dbern(psi[j])
  }
  # occupancy probabilities
  
  logit(psi[1:nsite]) <- alpha[1] + alpha[2] * stbathy[1:nsite]
  
  # detection matrix with obs for observations and state = true state
  # 1 for none species detected
  # 2 for species A
  # 3 for species B
  # 4 for both species detected
  
  ## observation process
  for(j in 1:nsite) {
    for(k in 1:nyear) {
      y[j, k] ~ dcat(obs[j, k, 1:4])
    }
  }
  
  # given state = unoccupied,
  for(j in 1:nsite) {
    for(k in 1:nyear) {
      # 
      obs[j, k, 1] <- 1 - z[j] + z[j] * (1 - ind.samm[j,k] * p_samm[j,k]) * (1 - ind.gd[j,k] * p_gdegem[j,k])  
      # prob obs = 1 no detection
      obs[j, k, 2] <- z[j] *(1 - ind.samm[j,k] * p_samm[j,k]) *  ind.gd[j,k] * p_gdegem[j,k]
      # prob obs = 2 detection by boat surveys only
      obs[j, k, 3] <- z[j] * (1 - ind.gd[j,k] * p_gdegem[j,k]) * ind.samm[j,k] * p_samm[j,k]
      # prob obs = 3 detection by boat surveys only
      obs[j, k, 4] <- z[j] * ind.gd[j,k] * ind.samm[j,k] * p_samm[j,k] * p_gdegem[j,k]
      # prob obs = 4 detection by both surveys
    }
  }
  
  # detection probabilities (pA function of pres/abs of B, as in Waddle et al 2010 page 1470)
  # VL: There are 2 detections probabilities now pA, pB
  
  for(j in 1:nsite) {
    for(k in 1:nyear) {
      # p_samm
      logit(p_samm[j,k])<- alpha.p_samm + beta[1] * eff.samm[j,k] 
      
      # p_samm
      logit(p_gdegem[j,k]) <- alpha.p_gdegem + beta[2] * eff.gd[j,k] 
    }
  }
  
  # priors for p
  alpha.p_samm ~ dnorm(0,1)
  alpha.p_gdegem ~ dnorm(0,1)
  for (i in 1:2){
    beta[i] ~ dnorm(0,1)
    alpha[i] ~ dnorm(0,1)
  }
})
```


Then, you choose to prepare either dolphin or trawlers dataset for the run, ie. run only one of the next paragraph.

### Dolphin

Get the ingredients for GAMs using package `jagam` developed by Simon Wood and basically hacks what is built by the package `mgcv`.

```{r message=FALSE, warning=FALSE}

unique(c(y))
dim(y)

yd <- matrix(0, nrow = dim(y)[1], ncol = dim(y)[2])
yd[y %in% c(2,4,10,12)] <- 2  # boat
yd[y %in% c(5,7,13,15)] <- 3 # aerial 
yd[y %in% c(6,8,14,16)] <- 4 # both

table(yd)
```

```{r eval = FALSE, include=FALSE}
# nimble
yd[yd ==0] <-  1
table(c(yd))
data <- list(y = yd, 
             eff.samm = seffS,
             ind.samm = effindS,
             eff.gd = seffG,
             ind.gd = effindG,
             stbathy= stbathy)

str(data)
constants <- list(nsite = dim(yd)[1], 
                  nyear = dim(yd)[2])

# Initial values for z {1:4} have to be consistent with the detections

#zinit <- matrix(NA, nrow = dim(yd)[1], ncol = dim(yd)[2])
zinit <- apply(yd,1, max)
zinit[zinit== 1] <- 0
zinit[zinit> 1] <- 1

# table
table(zinit)
table(c(yd))

inits <- list(z = zinit, 
              beta = rnorm(2, 0, 1), 
              alpha = rnorm(2,0,1),
              alpha.p_samm = rnorm(1,0,1),
              alpha.p_gdegem  = rnorm(1,0,1))
```


### Trawlers

```{r message=FALSE, warning=FALSE}

yt <- matrix(0, nrow = dim(y)[1], ncol = dim(y)[2])
yt[y %in% c(3,4,7,8)] <- 2  # boat
yt[y %in% c(9,10,14,13)] <- 3 # aerial 
yt[y %in% c(12,11,15,16)] <- 4 # both

table(yt)
```

```{r eval = FALSE, include=FALSE}
# nimble
yt[yt ==0] <-  1
table(c(yt))

dataT <- list(y = yt, 
              eff.samm = seffS,
              ind.samm = effindS,
              eff.gd = seffG,
              ind.gd = effindG,
              stbathy= stbathy)

str(dataT)
constantsT <- list(nsite = dim(yt)[1], 
                   nyear = dim(yt)[2])

# Initial values for z {1:4} have to be consistent with the detections

#zinit <- matrix(NA, nrow = dim(yd)[1], ncol = dim(yd)[2])
zinitT <- apply(yt,1, max)
zinitT[zinitT == 1] <- 0
zinitT[zinitT > 1] <- 1

# table
table(zinitT)
table(c(yt))

initsT <- list(z = zinitT, 
               beta = rnorm(2, 0, 1), 
               alpha = rnorm(2,0,1),
               alpha.p_samm = rnorm(1,0,1),
               alpha.p_gdegem  = rnorm(1,0,1))
```


## Build, compile and run model with NIMBLE

```{r eval = FALSE, include=TRUE}
Rmodel <- nimbleModel(intOcc, constantsT, dataT, initsT)
Rmodel$initializeInfo()
Rmodel$calculate() # T -662 D # - 401

conf <- configureMCMC(Rmodel)
conf$printMonitors() 

# Build and compile MCMC
Rmcmc <- buildMCMC(conf)
Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Cmodel)

# Run MCMC
t <- system.time(samplesTSG <- runMCMC(Cmcmc, niter = 150000, nburnin = 50000, nchains = 3,samplesAsCodaMCMC = TRUE))

```
